Too good to be false: Nonsignificant results revisited

### 2016-10-23 submission/manuscript.tex
Submission commit.

First, we compared the observed nonsignificant effect size distribution (computed with observed test results) to the expected nonsignificant effect size distribution under $H_0$. The expected effect size distribution under $H_0$ was approximated using simulation. We first randomly drew an observed test result (with replacement) and subsequently drew a random nonsignificant $p$-value between 0.05 and 1 (i.e., under the distribution of the $H_0$). Based on the drawn $p$-value and the degrees of freedom of the drawn test result, we computed the accompanying test statistic and the corresponding effect size (for details on effect size computation see Appendix B). This procedure was repeated 163,785 times, which is three times the number of observed nonsignificant test results (54,595). The collection of simulated results approximates the expected effect size distribution under $H_0$, assuming independence of test results. We inspected this possible dependency with the intra-class correlation ($ICC$), where $ICC=1$ indicates full dependency and $ICC=0$ indicates full independence. For the set of observed results, the ICC for $p$-values was .001, indicating independence of $p$-values within a paper. The resulting, expected effect size distribution was compared to the observed effect size distribution (i) across all journals and (ii) per journal. To test for differences between the expected and observed nonsignificant effect size distributions we applied the Kolmogorov-Smirnov test. This is a non-parametric goodness-of-fit test for equality of distributions, which is based on the maximum absolute deviation between the independent distributions being compared (denoted $D$; \cite{Massey1951-gj}).

@@

We also checked whether evidence of at least one false negative at the article level changed over time. Figure \ref{fig:fig4} depicts evidence across all articles per year, as a function of time (1985-2013); point size in the figure corresponds to the mean number of nonsignificant results per article (mean $k$) in that year. Interestingly, the proportion of articles with evidence for false negatives decreased from 77\% in 1985 to 55\% in 2013, despite the increase in mean $k$ (from 2.11 in 1985 to 0.55 in 2013). This decreasing evidence over time cannot be explained by a decrease in sample size over time, as sample size in psychology articles has stayed stable across time (see Figure \ref{fig:fig5}). One (partial) explanation of this surprising result is that in the early days researchers primarily reported fewer APA results and used to report relatively more APA results with 'marginally significant' $p$-values (i.e., $p$-values slightly larger than .05), compared to nowadays. This explanation is supported by both a smaller number of reported APA results in the past and the smaller mean reported nonsignificant $p$-value (0.222 in 1985, 0.386 in 2013). We do not know whether these marginally significant $p$-values were interpreted as evidence in favor of a finding (or not) and how these interpretations changed over time. Another potential explanation is that the effect sizes being studied have become smaller over time (mean effect $r$ 0.257 in 1985, 0.187 in 2013), which results in both higher $p$-values over time and lower power of the Fisher test. Using the data at hand, we cannot distinguish between the two explanations.

@@

The coding of the 178 results indicated that results rarely specify whether these are in line with the hypothesized effect (see Table \ref{tab:tab5}). For the 178 results, only 15 clearly stated whether their results were as expected, whereas the remaining 163 did not. Illustrative of the lack of clarity in expectations is the following quote: "\textit{As predicted, there was little gender difference [...] p < .06.}" As a result, the conditions significant-$H_0$ expected, nonsignificant-$H_0$ expected, and nonsignificant-$H_1$ expected contained too few results for meaningful investigation of evidential value (i.e., with sufficient statistical power).

\begin{table}[htbp]
\caption{Number of gender results coded per condition in a 2 (significance: significant or nonsignificant) by 3 (expectation: $H_0$ expected, $H_1$ expected, or no expectation) design. Cells printed in bold had sufficient results to inspect for evidential value.}
\rowcolors{2}{gray!25}{white}
% \begin{adjustbox}{width=\textwidth,totalheight=\textheight,keepaspectratio,angle=0}
\centering
\begin{tabular}{lrrr}
& $H_0$ expected & $H_1$ expected & No expectation \\
\hline
Significant    & 0           & \textbf{12}          & \textbf{76}             \\
Nonsignificant & 2           & 1           & \textbf{87}     \\
\hline
\end{tabular}
% \end{adjustbox}
\label{tab:tab5}
\end{table}

@@

The reanalysis of the nonsignificant RPP results using the Fisher method demonstrates that any conclusions on the validity of individual effects based on "failed" replications, as determined by statistical significance, is unwarranted and irresponsible. This was also noted by both the original RPP team \cite{Open_Science_Collaboration2015-zs,Anderson2016-bv} and in a critique of the RPP \cite{Gilbert2016-mi}. Replication efforts such as the RPP or the Many Labs project remove publication bias and result in a less biased assessment of the true effect size. Nonetheless, single replications should not be seen as the definitive result, considering that these results indicate there remains much uncertainty about whether a nonsignificant result is a true negative or a false negative. The explanation of this finding is that most of the RPP replications, although often statistically more powerful than the original studies, still did not have enough statistical power to distinguish a true small effect from a true zero effect \cite{Maxwell2015-yb}. Interpreting results of replications should therefore also take the precision of the estimate of both the original and replication into account \cite{Cumming2014-fi}.

@@

To conclude, the three applications that were presented indicate that false negatives remain a problem in the psychology literature, despite the decreased attention and that we should be wary to interpret statistically nonsignificant results as there being no effect in reality. One way to combat this interpretation of statistically nonsignificant results is to incorporate testing for potential false negatives, which the Fisher method facilitates in a highly approachable manner  (a spreadsheet for carrying out such a test is available at \url{https://osf.io/tk57v/}).

@@

We simulated false negative $p$-values according to the following six steps (see Figure \ref{fig:appendixa}). First, we determined the critical value under the null distribution. Second, we determined the alternative distribution by computing the non-centrality parameter ($\delta=(\eta^2/1-\eta^2)N$; \cite{Steiger1997-qq,Smithson2001-aw}). Third, we calculated the probability that a result under the alternative distribution was, in fact, nonsignificant (i.e., $\beta$). Fourth, we randomly sampled, uniformly, a value between $0-\beta$. Fifth, with this value we determined the accompanying $t$-value. Finally, we computed the $p$-value for this $t$-value under the null distribution.

>>>>

First, we compared the observed nonsignificant effect size distribution (computed with observed test results) to the expected nonsignificant effect size distribution under $H_0$. The expected effect size distribution under $H_0$ was approximated using simulation. We first randomly drew an observed test result (with replacement) and subsequently drew a random nonsignificant $p$-value between 0.05 and 1 (i.e., under the distribution of the $H_0$). Based on the drawn $p$-value and the degrees of freedom of the drawn test result, we computed the accompanying test statistic and the corresponding effect size (for details on effect size computation see Appendix B). This procedure was repeated 163,785 times, which is three times the number of observed nonsignificant test results (54,595). The collection of simulated results approximates the expected effect size distribution under $H_0$, assuming independence of test results. We inspected this possible dependency with the intra-class correlation ($ICC$), where $ICC=1$ indicates full dependency and $ICC=0$ indicates full independence. For the set of observed results, the ICC for nonsignificant $p$-values was 0.001, indicating independence of $p$-values within a paper. The resulting, expected effect size distribution was compared to the observed effect size distribution (i) across all journals and (ii) per journal. To test for differences between the expected and observed nonsignificant effect size distributions we applied the Kolmogorov-Smirnov test. This is a non-parametric goodness-of-fit test for equality of distributions, which is based on the maximum absolute deviation between the independent distributions being compared (denoted $D$; \cite{Massey1951-gj}).

@@

We also checked whether evidence of at least one false negative at the article level changed over time. Figure \ref{fig:fig4} depicts evidence across all articles per year, as a function of year (1985-2013); point size in the figure corresponds to the mean number of nonsignificant results per article (mean $k$) in that year. Interestingly, the proportion of articles with evidence for false negatives decreased from 77\% in 1985 to 55\% in 2013, despite the increase in mean $k$ (from 2.11 in 1985 to 4.52 in 2013). This decreasing proportion of papers with evidence over time cannot be explained by a decrease in sample size over time, as sample size in psychology articles has stayed stable across time (see Figure \ref{fig:fig5}). One (at least partial) explanation of this surprising result is that in the early days researchers primarily reported fewer APA results and used to report relatively more APA results with 'marginally significant' $p$-values (i.e., $p$-values slightly larger than .05), compared to nowadays. This explanation is supported by both a smaller number of reported APA results in the past and the smaller mean reported nonsignificant $p$-value (0.222 in 1985, 0.386 in 2013). We do not know whether these marginally significant $p$-values were interpreted as evidence in favor of a finding (or not) and how these interpretations changed over time. Another potential explanation is that the effect sizes being studied have become smaller over time (mean effect $r$ 0.257 in 1985, 0.187 in 2013), which results in both higher $p$-values over time and lower power of the Fisher test. Using the data at hand, we cannot distinguish between the two explanations.

@@

The coding of the 178 results indicated that results rarely specify whether these are in line with the hypothesized effect (see Table \ref{tab:tab5}). For the 178 results, only 15 clearly stated whether their results were as expected, whereas the remaining 163 did not. Illustrative of the lack of clarity in expectations is the following quote: "\textit{As predicted, there was little gender difference [...] p < .06.}" There were two results that were presented as significant but contained $p$-values larger than .05; these two were dropped (i.e., 176 results were analyzed). As a result, the conditions significant-$H_0$ expected, nonsignificant-$H_0$ expected, and nonsignificant-$H_1$ expected contained too few results for meaningful investigation of evidential value (i.e., with sufficient statistical power).

\begin{table}[htbp]
\caption{Number of gender results coded per condition in a 2 (significance: significant or nonsignificant) by 3 (expectation: $H_0$ expected, $H_1$ expected, or no expectation) design. Cells printed in bold had sufficient results to inspect for evidential value.}
\rowcolors{2}{gray!25}{white}
% \begin{adjustbox}{width=\textwidth,totalheight=\textheight,keepaspectratio,angle=0}
\centering
\begin{tabular}{lrrr}
& $H_0$ expected & $H_1$ expected & No expectation \\
\hline
Significant    & 0           & \textbf{11}          & \textbf{75}             \\
Nonsignificant & 2           & 1           & \textbf{87}     \\
\hline
\end{tabular}
% \end{adjustbox}
\label{tab:tab5}
\end{table}

@@

The reanalysis of the nonsignificant RPP results using the Fisher method demonstrates that any conclusions on the validity of individual effects based on "failed" replications, as determined by statistical significance, is unwarranted and irresponsible. This was also noted by both the original RPP team \cite{Open_Science_Collaboration2015-zs,Anderson2016-bv} and in a critique of the RPP \cite{Gilbert2016-mi}. Replication efforts such as the RPP or the Many Labs project remove publication bias and result in a less biased assessment of the true effect size. Nonetheless, single replications should not be seen as the definitive result, considering that these results indicate there remains much uncertainty about whether a nonsignificant result is a true negative or a false negative. The explanation of this finding is that most of the RPP replications, although often statistically more powerful than the original studies, still did not have enough statistical power to distinguish a true small effect from a true zero effect \cite{Maxwell2015-yb}. Interpreting results of replications should therefore also take the precision of the estimate of both the original and replication into account \cite{Cumming2014-fi} and publication bias of the original studies \cite{10.1371/journal.pone.0149794}.

@@

To conclude, our three applications indicate that false negatives remain a problem in the psychology literature, despite the decreased attention and that we should be wary to interpret statistically nonsignificant results as there being no effect in reality. One way to combat this interpretation of statistically nonsignificant results is to incorporate testing for potential false negatives, which the Fisher method facilitates in a highly approachable manner (a spreadsheet for carrying out such a test is available at \url{https://osf.io/tk57v/}).

@@

We simulated false negative $p$-values according to the following six steps (see Figure \ref{fig:appendixa}). First, we determined the critical value under the null distribution. Second, we determined the distribution under the alternative hypothesis by computing the non-centrality parameter ($\delta=(\eta^2/1-\eta^2)N$; \cite{Steiger1997-qq,Smithson2001-aw}). Third, we calculated the probability that a result under the alternative hypothesis was, in fact, nonsignificant (i.e., $\beta$). Fourth, we randomly sampled, uniformly, a value between $0-\beta$. Fifth, with this value we determined the accompanying $t$-value. Finally, we computed the $p$-value for this $t$-value under the null distribution.
